How to \emph{not} write Virtual Machines for Dynamic Languages

### 2007-05-27 talk/dyla2007/dyla.tex
Increase the skip.  Seems to have no visible effect when Carl Friedrich compiles the pdf...
53% of a review

We argue in this paper that one should not write interpreters for dynamic
languages manually but rather use meta-programming techniques and raise the
overall level at which they are are implemented. We believe this to be
ultimately a better investment of efforts than the development of more advanced
general-purpose object oriented VMs.

\smallskip

Dynamic languages are traditionally implemented by writing a virtual machine for
them, centered around an interpreter and/or a built-in compiler and providing
the object model and memory management. When a language becomes more successful,
the limitations of such an implementation lead to the emergence of alternative
implementations that try to solve some of the problems. Another reason for new
implementations is the desire to have the language integrate well with existing,
well-tuned object-oriented virtual machine like Java's. In this paper, we
describe the mechanisms that lead to an abundance of implementations, and
explore some of the limitations of standard VMs.  We propose a different
complementary alternative to writing VMs by hand and dealing with low-level
details, validated by the PyPy project: flexibly generating virtual machines
from a single abstract language ``specification'', inserting features and
low-level details automatically -- including good just-in-time compilers tuned
to the dynamic language at hand.
\footnote{This research was partially supported by the EU funded
 project: IST 004779 PyPy (PyPy: Implementing Python in Python).} \\

@@

The efforts required to build a new virtual machine are relatively
important.  This is particularly true for languages which are complex
and in constant evolution. Language implementation communities from an
open-source or academic context have only limited resources. Therefore they
cannot afford to have a highly complex implementation and often chose simpler
techniques even if that entails lower execution speed. Similarly fragmentation
(for example because of other implementations of the same language) is a
problem because it divides available resources. All these points also apply to
the implementation of domain-specific languages where it is important to keep
the implementation effort small.

For these reasons writing a virtual machine in C is problematic because it
forces the language implementer to deal with many low-level details (like
garbage collection and threading issues). Limitations
of the C implementation lead to alternative implementations which draw
work-power from the reference implementation. An alternative to writing
implementations in C is to build them on top of one of the newer object oriented
virtual machines (``OO VM'') such as the JVM or the CLR. This is often wanted by
the community anyway, since it leads to better re-usability of libraries of
these platforms. However, if a C implementation existed before the
implementation of such a VM is started, this enters in conflict with the goal of
having to maintain essentially a single, simple enough implementation for a
given programming language: as the language becomes popular, there will be a
demand for having it run on various platforms -- high-level VMs as well as
C-level environments.

@@

With the emergence of .NET and the JVM as interesting language
implementation platforms, an argument that is sometimes made is that
communities should only develop an implementation of their language for
one of these platforms.\footnote{Preferably the argument author's favourite
one.}

@@

Implementing a language on top of an existing OO VM is in many ways easier than
implementing it in C. Let's take a look at the advantages that are usually
cited for basing a
language implementation of a dynamic language on a standard object oriented
virtual machine, for example the JVM or the CLR. The central theme of the
benefits  of OO VMs is the ability to implement certain hard things only once
and share the benefits between all language implementations on top of the OO VM.

@@

\item
\emph{Ease of implementation:} The implementation of a language on top of an OO
VM is easier because it starts at a higher level than C. Usually a
high-level language like Java or C\# is used for the language implementation,
which both offer the language implementer a much higher level of abstraction
than when implementing in C.

@@

\noindent
At a closer look, some of these advantages are only partially true in practice:

@@

The described proliferation of language implementations is a big problem for
language communities. Although most individual implementations exist for good
reasons, the sum of all of them and the need to keep them synchronized with the
reference implementations lead to a lot of duplicated work and division of
efforts. This is especially true for open source languages which tend to evolve
quickly. At any one point in time some of the implementations will lag behind
which makes writing code which can work on all of the implementations harder.

@@

In the next section we are exploring a different approach for implementing
dynamic languages that we hope is able to solve many of the problems of
implementing a language, in particular the problem of an explosion of the number
of implementations.

>>>>

We argue in this paper that one should not write interpreters for dynamic
languages manually but rather use meta-programming techniques and raise the
overall level at which they are are implemented. We believe this to be
ultimately a better investment of efforts than the development of more and more advanced
general-purpose object oriented VMs.

\medskip

Dynamic languages are traditionally implemented by writing a virtual machine for
them, centered around an interpreter and/or a built-in compiler and providing
the object model and memory management. When a language becomes more popular,
the limitations of such an implementation lead to the emergence of alternative
implementations that try to solve some of the problems. Another reason for new
implementations is the desire to have the language integrate well with existing,
well-tuned object-oriented virtual machine like the Java Virtual Machine. In this paper, we
describe the mechanisms that lead to an abundance of implementations, and
explore some of the limitations of standard VMs.  We propose a different
complementary alternative to writing VMs by hand and dealing with low-level
details, validated by the PyPy project: flexibly generating virtual machines
from a single abstract language ``specification'', inserting features and
low-level details automatically -- including good just-in-time compilers tuned
to the dynamic language at hand.
\footnote{This research was partially supported by the EU funded
 project: IST 004779 PyPy (PyPy: Implementing Python in Python).} \\

@@

The effort required to build a new virtual machine are relatively
large.  This is particularly true for languages which are complex
and in constant evolution. Language implementation communities from an
open-source or academic context have only limited resources. Therefore they
cannot afford to have a highly complex implementation and often chose simpler
techniques even if that entails lower execution speed. Similarly fragmentation
(for example because of other implementations of the same language) is a
problem because it divides available resources. All these points also apply to
the implementation of domain-specific languages where it is important to keep
the implementation effort small.

For these reasons writing a virtual machine in C is problematic because it
forces the language implementer to deal with many low-level details (like
garbage collection and threading issues). Limitations
of the C implementation lead to alternative implementations which draw
work-power from the reference implementation. An alternative to writing
implementations in C is to build them on top of one of the newer object oriented
virtual machines (``OO VM'') such as the JVM or the CLR. This is often wanted by
the community anyway, since it leads to the ability to re-use the libraries of
these platforms. However, if a C implementation existed before the
implementation of such a VM is started, this enters into conflict with the goal of
having to maintain essentially a single, simple enough implementation for a
given programming language: as the language becomes popular, there will be a
demand for having it run on various platforms -- high-level VMs as well as
C-level environments.

@@

With the emergence of .NET and the JVM as interesting language
implementation platforms, an argument that is sometimes made is that
communities should only develop an implementation of their language
for one of these platforms (preferably the argument author's favourite
one).

@@

Implementing a language on top of an existing OO VM is in many ways easier than
implementing it in C. Let's take a look at the advantages that are usually
cited for basing a
language implementation of a dynamic language on a standard object oriented
virtual machine, for example the JVM or the CLR.

@@

\item
\emph{Ease of implementation:} The implementation of a language on top of an OO
VM is easier because it starts at a higher level than C. Usually a
high-level language like Java or C\# is used for the language implementation,
both of which offer the language implementer a much higher level of abstraction
than when implementing in C.

@@

\noindent
The central theme of the benefits of OO VMs is the ability to
implement certain hard things only once and share the benefits between
all language implementations on top of the OO VM.  At a closer look,
some of these advantages are only partially true in practice:

@@

The described proliferation of language implementations is a big problem for
language communities. Although most individual implementations exist for good
reasons, the sum of all of them and the need to keep them synchronized with the
reference implementations lead to a lot of duplicated work and division of
effort. This is especially true for open source languages which tend to evolve
quickly. At any one point in time some of the implementations will lag behind
which makes writing code which can work on all of the implementations harder.

@@

In the next section we explore a different approach for implementing
dynamic languages that we hope is able to solve many of the problems of
implementing a language, in particular the problem of an explosion of the number
of implementations.
